---
title: "Logic in R"
author: "Colin Rundel"
date: "2019-01-15"
output:
  xaringan::moon_reader:
    css: "slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

exclude: true

```{r, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width=80
)

htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

---
class: middle
count: false

# (Almost) <br/> Everything is a Vector

---

## Vectors

The fundamental building block of data in R are vectors (collections of related values, objects, other data structures, etc).

--
<br/>

R has two types of vectors:

* **atomic** vectors 

    - homogeneous collections of the *same* type (e.g. all logical values, all numbers, or all character strings).

* **generic** vectors
  
    - heterogeneous collections of *any* type of R object, even other lists (meaning they can have a hierarchical/tree-like structure).

---

## Atomic Vectors

R has six atomic vector types: 
  
<br/>
  
.center[
`logical`,  `double`, `integer`, `character`, `complex`, `raw`      
]

<br/>

For today we'll mostly worry about the first type, we'll discuss the remainder over the next couple of lextures (note that the last two almost never come up).

---
count: false
class: middle

# Conditionals

---

## Logical (boolean) operators

|  Operator                     |  Operation    |  Vectorized? 
|:-----------------------------:|:-------------:|:------------:
| <code>x &#124; y</code>       |  or           |   Yes        
| `x & y`                       |  and          |   Yes        
| `!x`                          |  not          |   Yes        
| <code>x &#124;&#124; y</code> |  or           |   No         
| `x && y`                      |  and          |   No         
|`xor(x,y)`                     |  exclusive or |   Yes        


---

## Vectorized?

```{r}
x = c(TRUE,FALSE,TRUE)
y = c(FALSE,TRUE,TRUE)
```

.pull-left[
```{r}
x | y
x || y
```
]

.pull-right[
```{r}
x & y
x && y
```
]

---

## Vectorization and arithmatic

Almost all of the basic mathematical operations (and many other functions) in R are vectorized as well.

.pull-left[
```{r}
c(1,2,3) + c(3,2,1)
c(1,2,3) / c(3,2,1)
```
]

.pull-right[
```{r}
log(c(1, 3, 0))
sin(c(1,2,3))
```
]



---

## Length coercion

```{r}
x = c(TRUE,FALSE,TRUE)
y = c(TRUE)
z = c(FALSE,TRUE)
```

--

.pull-left[
```{r}
x | y
x & y
```
]

--

.pull-right[
```{r}
y | z
y & z
```
]

--

```{r}
x | z
```

---

## Comparisons

  Operator  |  Comparison                |  Vectorized?
:----------:|:--------------------------:|:----------------:
 `x < y`    |  less than                 |  Yes
 `x > y`    |  greater than              |  Yes
 `x <= y`   |  less than or equal to     |  Yes
 `x >= y`   |  greater than or equal to  |  Yes
 `x != y`   |  not equal to              |  Yes
 `x == y`   |  equal to                  |  Yes
 `x %in% y` |  contains                  |  Yes (over `x`)


---

## Comparisons

```{r}
x = c("A","B","C")
z = c("A")
```

.pull-left[
```{r}
x == z
x != z
x > z
```
]

--

.pull-right[
```{r}
x %in% z
z %in% x
```
]
 

---

## Conditional Control Flow

Conditional execution of code blocks is achieved via `if` statements. 

```{r}
x = c(1,3)
```

--

```{r}
if (3 %in% x)
  print("This!")
```

--

```{r}
if (1 %in% x)
  print("That!")
```

--

```{r}
if (5 %in% x)
  print("Other!")
```


---

## `if` is not vectorized

```{r}
x = c(1,3)
```

--

```{r}
if (x %in% 3)
  print("Now Here!")
```

--

```{r}
if (x %in% 1)
  print("Now Here!")
```


---

## Collapsing logical vectors

There are a couple of helper functions for collapsing a logical vector down to a single value: `any`, `all`

```{r}
x = c(3,4,1)
```

.pull-left[
```{r}
x >= 2
any(x >= 2)
all(x >= 2)
```
]

.pull-right[
```{r}
x <= 4
any(x <= 4)
all(x <= 4)
```
]


---

## Nesting Conditionals

.pull-left[
```{r}
x = 3
if (x < 0) {
  "Negative"
} else if (x > 0) {
  "Positive"
} else {
  "Zero"
}
```
]

.pull-right[
```{r}
x = 0
if (x < 0) {
  "Negative"
} else if (x > 0) {
  "Positive"
} else {
  "Zero"
}
```
]

---
class: middle
count: false

# Error Checking

---

## `stop` and `stopifnot`

Often we want to validate user input or function arguments - if our assumptions are not met then we often want to report the error and stop execution. 

```{r error=TRUE}
ok = FALSE
if (!ok)
  stop("Things are not ok.")

stopifnot(ok)
```

*Note* - an error (like the one generated by `stop`) will prevent an RMarkdown document from compiling unless `error=TRUE` is set for that code chunk

---

## Style choices


.pull-left[
Do stuff:
```{r eval=FALSE}
if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
} else if (condition_error) {
  stop("Condition error occured")
}
```
]

.pull-right[
Do stuff (better):
```{r eval=FALSE}
# Do stuff better
if (condition_error) {
  stop("Condition error occured")
}

if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
}
```
]

---

## Exercise 1

Write a set of conditional(s) that satisfies the following requirements,

* If `x` is greater than 3 and `y` is less than or equal to 3 then print "Hello world!"

* Otherwise if `x` is greater than 3 print "!dlrow olleH"

* If `x` is less than or equal to 3 then print "Something else ..."

* Stop execution if x is odd and y is even and report an error, don't print any of the text strings above.


Test out your code by trying various values of `x` and `y`.


---
class: middle
count: false

# Loops

---

## `for` loops

Simplest, and most common type of loop in R - given a vector iterate through the elements and evaluate the code block for each.


```{r}
res = c()
for(x in 1:10) {
  res = c(res, x^2)
}
res
```

--

```{r}
res = c()
for(y in list(1:3, LETTERS[1:7], c(TRUE,FALSE))) {
  res = c(res, length(y))
}
res
```

<br/>

*Note* - the code above is terrible for several reasons, you should never write anything that looks like this

---

## `while` loops

Repeat until the given condition is **not** met (i.e. evaluates to `FALSE`)

```{r}
i = 1
res = rep(NA,10)

while (i <= 10) {
  res[i] = i^2
  i = i+1
}

res
```

---

## `repeat` loops

Repeat until `break`

```{r}
i = 1
res = rep(NA,10)

repeat {
  res[i] = i^2
  i = i+1
  if (i > 10)
    break
}

res
```

---
class: split-50

## Special keywords - `break` and `next`

These are special actions that only work *inside* of a loop

* `break` - ends the current *loop* (inner-most)
* `next` - ends the current *iteration*

.pull-left[
```{r}
res = c()
for(i in 1:10) {
    if (i %% 2 == 0)
        break
    res = c(res, i)
    print(res)
}
```
]

.pull-right[
```{r}
res = c()
for(i in 1:10) {
    if (i %% 2 == 0)
        next
    res = c(res,i)
    print(res)
}
```
]


---

## Some helper functions

Often we want to use a loop across the indexes of an object and not the elements themselves. There are several useful functions to help you do this: `:`, `length`, `seq`, `seq_along`, `seq_len`, etc.

.pull-left[
```{r}
4:7
length(4:7)
seq(4,7)
```
]

.pull-right[
```{r}
seq_along(4:7)
seq_len(length(4:7))
seq(4,7,by=2)
```
]

---

## Exercise 2

Below is a vector containing all prime numbers between 2 and 100:

.center[
```r
primes = c( 2,  3,  5,  7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 
      43, 47, 53, 59, 61, 67, 71, 73, 79, 83, 89, 97)
```
]

If you were given the vector `x = c(3,4,12,19,23,51,61,63,78)`, write the R code necessary to print only the values of `x` that are *not* prime (without using subsetting or the `%in%` operator). 

Your code should use *nested* loops to iterate through the vector of primes and `x`.




---
count: false

# Acknowledgments

Above materials are derived in part from the following sources:

* Hadley Wickham - [Advanced R](http://adv-r.had.co.nz/)
* [R Language Definition](http://stat.ethz.ch/R-manual/R-devel/doc/manual/R-lang.html)


