---
title: dbplyr
author: "Colin Rundel"
date: "2019-04-04"
output:
  xaringan::moon_reader:
    css: "slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
exclude: true

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width = 80,
  tibble.width = 80
)

knitr::opts_chunk$set(
  fig.align = "center"
)

htmltools::tagList(rmarkdown::html_dependency_font_awesome())

library(dplyr)
```

---

## Database backends

```{r}
library(dplyr)

ls("package:dplyr") %>% purrr::keep(stringr::str_detect, "^src")

?src_sqlite
```

---

## Creating a database 

.small[
```{r eval=FALSE}
(db = dplyr::src_sqlite("/data/nyc_parking/nyc_parking_2014_cleaned.sqlite", create = TRUE))

##  src:  sqlite 3.22.0 [/data/nyc_parking/nyc_parking_2014_cleaned.sqlite]
##  tbls:
```
]

--

.small[
```{r eval=FALSE}
nyc_sql = dplyr::copy_to(db, nyc, temporary = FALSE)
db

## src:  sqlite 3.22.0 [/data/nyc_parking/nyc_parking_2014_cleaned.sqlite]
## tbls: nyc, sqlite_stat1, sqlite_stat4
```
]

--

.small[
```r
pryr::object_size(db)
## 6.54 kB

pryr::object_size(nyc_sql)
## 9.54 kB
```

```shell
> ls -lah /data/nyc_parking/*.sqlite

-rw-r--r-- 1 cr173 visitor 698M Nov  5 13:57 /data/nyc_parking/nyc_parking_2014_cleaned.sqlite
```
]

---

.small[
```{r eval=FALSE}
nyc_sql = dplyr::tbl(db,"nyc")

class(nyc_sql)
## [1] "tbl_dbi"  "tbl_sql"  "tbl_lazy" "tbl" 
```
]

--

.small[
```r
nyc_sql
## # Source:   table<nyc> [?? x 18]
## # Database: sqlite 3.22.0 [/data/nyc_parking/nyc_parking_2014_cleaned.sqlite]
##    registration_state Plate_Type issue_date Violation_Code Vehicle_Body_Type Vehicle_Make issuing_agency
##                 <chr>      <chr>      <dbl>          <int>             <chr>        <chr>          <chr>
##  1                 NJ        PAS      15706             20              SUBN        CHRYS              P
##  2                 MI        PAS      15706             99               SDN        TOYOT              P
##  3                 NY        PAS      15706             20              SUBN         FORD              P
##  4                 NY        PAS      15706             20               SDN        INFIN              P
##  5                 NY        SRF      15706             20               SDN        NISSA              P
##  6                 NY        PAS      15706             46              SUBN        CHRYS              P
##  7                 NY        PAS      15706             62               SDN        HYUND              P
##  8                 NY        PAS      15706             98               SDN        TOYOT              P
##  9                 NY        PAS      15706             71               SDN        TOYOT              P
## 10                 NY        PAS      15706             71              SUBN         FORD              P
## # ... with more rows, and 11 more variables: violation_location <chr>, violation_precinct <int>,
## #   violation_time <chr>, house_number <chr>, street_name <chr>, intersecting_street <chr>,
## #   vehicle_color <chr>, issue_day <int>, issue_month <dbl>, issue_year <dbl>, issue_wday <chr>
```
]

--

.small[
```r
str(nyc_sql)
## List of 2
##  $ src:List of 2
##   ..$ con  :Formal class 'SQLiteConnection' [package "RSQLite"] with 7 slots
##   .. .. ..@ ptr                :<externalptr> 
##   .. .. ..@ dbname             : chr "/data/nyc_parking/nyc_parking_2014_cleaned.sqlite"
##   .. .. ..@ loadable.extensions: logi TRUE
##   .. .. ..@ flags              : int 70
##   .. .. ..@ vfs                : chr ""
##   .. .. ..@ ref                :<environment: 0x55788ecb6f38> 
##   .. .. ..@ bigint             : chr "integer64"
##   ..$ disco:<environment: 0x557888dd9840> 
##   ..- attr(*, "class")= chr [1:3] "src_dbi" "src_sql" "src"
##  $ ops:List of 2
##   ..$ x   : 'ident' chr "nyc"
##   ..$ vars: chr [1:18] "registration_state" "plate_type" "issue_date" "violation_code" ...
##   ..- attr(*, "class")= chr [1:3] "op_base_remote" "op_base" "op"
##  - attr(*, "class")= chr [1:4] "tbl_dbi" "tbl_sql" "tbl_lazy" "tbl"
```
]

---

## Using dplyr with sqlite 

.small[
```{r, eval=FALSE}
(addr = nyc_sql %>%
        select(issue_date, issuing_agency, violation_precinct, house_number, street_name) %>%
        filter(violation_precinct >=1, violation_precinct <= 34)
)

## # Source:   lazy query [?? x 5]
## # Database: sqlite 3.22.0 [/data/nyc_parking/nyc_parking_2014_cleaned.sqlite]
##    issue_date issuing_agency violation_precinct house_number street_name    
##         <dbl> <chr>                       <int> <chr>        <chr>          
##  1      15706 P                              28 102          W 123 ST       
##  2      15706 P                              23 2121         1 AVE          
##  3      15706 P                              23 60           E 106 ST       
##  4      15706 P                               5 54           ELIZABETH ST   
##  5      15706 P                              26 488-490      ST NICHOLAS AVE
##  6      15706 P                              26 1420         AMSTERDAM AVE  
##  7      15706 P                              25 219          E 121 ST       
##  8      15706 P                              12 630          LEXINGTON AVE  
##  9      15706 P                              18 413          48 TH ST       
## 10      15706 P                              25 2123         MADISON AVE    
## # ... with more rows
```
]

---

## SQL Query

.small[
```{r, eval=FALSE}
class(addr)

## [1] "tbl_dbi"  "tbl_sql"  "tbl_lazy" "tbl" 

show_query(addr)

## <SQL>
## SELECT *
## FROM (SELECT `issue_date`, `issuing_agency`, `violation_precinct`, `house_number`, `street_name`
## FROM `nyc`)
## WHERE ((`violation_precinct` >= 1.0) AND (`violation_precinct` <= 34.0))
```
]


---

## SQL Grouping 

```{r, eval=FALSE}
addr %>% group_by(issuing_agency, violation_precinct) %>% summarize(n=n())

##  # Source:   lazy query [?? x 3]
##  # Database: sqlite 3.22.0 [/data/nyc_parking/nyc_parking_2014_cleaned.sqlite]
##  # Groups:   issuing_agency
##     issuing_agency violation_precinct     n
##              <chr>              <int> <int>
##   1              A                  1    13
##   2              A                  7     1
##   3              A                 10    24
##   4              A                 11     1
##   5              A                 14    47
##   6              A                 33    11
##   7              B                 25     2
##   8              C                  5    73
##   9              C                 13     7
##  10              D                  1     1
##  # ... with more rows
```

---

## SQL Query

```{r, eval=FALSE}
addr %>% group_by(issuing_agency, violation_precinct) %>% summarize(n=n()) %>% show_query()

## <SQL>
## SELECT `issuing_agency`, `violation_precinct`, COUNT() AS `n`
## FROM (SELECT `issue_date` AS `issue_date`, `issuing_agency` AS `issuing_agency`, `violation_precinct` AS `## violation_precinct`, `house_number` AS `house_number`, `street_name` AS `street_name`
## FROM `nyc`)
## WHERE ((`violation_precinct` >= 1.0) AND (`violation_precinct` <= 34.0))
## GROUP BY `issuing_agency`, `violation_precinct`
```

---

## SQL Translation 

.small[
In general, dplyr / dbplyr knows how to translate basic math, logical, and summary functions from R to SQL.
dbplyr has a function, `translate_sql`, that lets you experiment with how R functions are translated to SQL. 

```{r, error=TRUE}
dbplyr::translate_sql(x == 1 & (y < 2 | z > 3))
dbplyr::translate_sql(x ^ 2 < 10)
dbplyr::translate_sql(x %% 2 == 10)
```

```{r error=TRUE}
dbplyr::translate_sql(paste(x,y))
dbplyr::translate_sql(mean(x))
```
]

---

.small[
```{r error=TRUE}
dbplyr::translate_sql(sd(x))
dbplyr::translate_sql(paste(x,y))
dbplyr::translate_sql(cumsum(x))
dbplyr::translate_sql(lag(x))
dbplyr::translate_sql(lm(y~x))
```
]


---

## Complications

.small[
```{r, eval=FALSE}
addr %>% mutate(address = paste(house_number, street_name))

## # Source:   lazy query [?? x 6]
## # Database: sqlite 3.22.0 [/data/nyc_parking/nyc_parking_2014_cleaned.sqlite]
##    issue_date issuing_agency violation_precinct house_number street_name     address                
##         <dbl> <chr>                       <int> <chr>        <chr>           <chr>                  
##  1      15706 P                              28 102          W 123 ST        102 W 123 ST           
##  2      15706 P                              23 2121         1 AVE           2121 1 AVE             
##  3      15706 P                              23 60           E 106 ST        60 E 106 ST            
##  4      15706 P                               5 54           ELIZABETH ST    54 ELIZABETH ST        
##  5      15706 P                              26 488-490      ST NICHOLAS AVE 488-490 ST NICHOLAS AVE
##  6      15706 P                              26 1420         AMSTERDAM AVE   1420 AMSTERDAM AVE     
##  7      15706 P                              25 219          E 121 ST        219 E 121 ST           
##  8      15706 P                              12 630          LEXINGTON AVE   630 LEXINGTON AVE      
##  9      15706 P                              18 413          48 TH ST        413 48 TH ST           
## 10      15706 P                              25 2123         MADISON AVE     2123 MADISON AVE       
## # ... with more rows

addr %>% mutate(address = paste(house_number, street_name)) %>% show_query()

## <SQL>
## SELECT `issue_date`, `issuing_agency`, `violation_precinct`, `house_number`, `street_name`, 
##        `house_number` || ' ' || ## `street_name` AS `address`
## FROM (SELECT `issue_date`, `issuing_agency`, `violation_precinct`, `house_number`, `street_name`
##       FROM `nyc`)
## WHERE ((`violation_precinct` >= 1.0) AND (`violation_precinct` <= 34.0))
```
]

---

## (Unfair) Timings 

.small[
```{r, eval=FALSE}
system.time(
  nyc %>%
    select(issue_date, issuing_agency, violation_precinct, house_number, street_name) %>%
    filter(violation_precinct >=1, violation_precinct <= 34) %>%
    group_by(issuing_agency, violation_precinct) %>%
    summarize(n=n())
)

##   user  system elapsed 
##  0.639   0.099   0.740 
```
]

.small[
```{r, eval=FALSE}
system.time(
  nyc_sql %>%
    select(issue_date, issuing_agency, violation_precinct, house_number, street_name) %>%
    filter(violation_precinct >=1, violation_precinct <= 34) %>%
    group_by(issuing_agency, violation_precinct) %>%
    summarize(n=n())
)

##   user  system elapsed 
##  0.024   0.011   0.034 
```
]

--

`nyc_sql` was ~22x times faster than `nyc`, but the former is disk based while the latter is in memory, why this discrepancy?

---

## Laziness

dplyr / dbplyr uses lazy evaluation as much as possible, particularly when working with non-local backends.

* When building a query, we don't want the entire table, often we want just enough to check if our query is working.

* Since we would prefer to run one complex query over many simple queries, laziness allows for verbs to be strung together.

* Therefore, by default `dplyr`
    
    * won't connect and query the database until absolutely necessary (e.g. show output),

    * and unless explicitly told to, will only query a handful of rows to give a sense of what the result will look like


---

.small[
```{r eval=FALSE}
nyc_sql %>%
  select(issue_date, issuing_agency, violation_precinct, house_number, street_name) %>%
  filter(violation_precinct >=1, violation_precinct <= 34) %>%
  group_by(issuing_agency, violation_precinct) %>%
  summarize(n=n())

## # Source:   lazy query [?? x 3]
## # Database: sqlite 3.22.0 [/data/nyc_parking/nyc_parking_2014_cleaned.sqlite]
## # Groups:   issuing_agency
##    issuing_agency violation_precinct     n
##             <chr>              <int> <int>
##  1              A                  1    13
##  2              A                  7     1
##  3              A                 10    24
##  4              A                 11     1
##  5              A                 14    47
##  6              A                 33    11
##  7              B                 25     2
##  8              C                  5    73
##  9              C                 13     7
## 10              D                  1     1
## # ... with more rows
```
]

---

## Full query 
.small[

To force a full query *and* a return of the complete result it is necessart to use the `collect` function.

```{r, eval=FALSE}
system.time({
  nyc_sql %>%
    select(issue_date, issuing_agency, violation_precinct, house_number, street_name) %>%
    filter(violation_precinct >=1, violation_precinct <= 34) %>%
    group_by(issuing_agency, violation_precinct) %>%
    summarize(n=n()) %>%
    collect()
})

##   user  system elapsed 
##  5.915   0.507   6.445 

## # A tibble: 199 x 3
## # Groups:   issuing_agency [15]
##    issuing_agency violation_precinct     n
##             <chr>              <int> <int>
##  1              A                  1    13
##  2              A                  7     1
##  3              A                 10    24
##  4              A                 11     1
##  5              A                 14    47
##  6              A                 33    11
##  7              B                 25     2
##  8              C                  5    73
##  9              C                 13     7
## 10              D                  1     1
## # ... with 189 more rows
```


`compute` and `collapse` also force a full query but have slightly different behavior and return types. 

]

---

## Acknowledgments

Above materials are derived in part from the following sources:

* [dbplyr - Introduction Vignette](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html)
* [dbplyr - SQL Translation](https://cran.r-project.org/web/packages/dbplyr/vignettes/sql-translation.html)